Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-10-28 paper/manuscript.tex
references for solvation

@@

The standard molar Gibbs free energy of the solvation of a gas species $A$ in a solvent $v$
\begin{equation}
A(\mathrm{g})  \leftrightharpoons A(v)
\end{equation}
is
\begin{equation}
  \Delta G^\circ_{\mathrm{slv}} = \mu^\circ_{A(v)}  - \mu^\circ_{A(g)} = -k_B T \ln \frac{P^\circ}{C^\circ k_B T} - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}}
  \label{eq:DG0-solvation-def}
\end{equation}
where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel. The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases. The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below). Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.

However, equilibrium experimental gas solvation data is most often available in terms of Henry's constants $k_H$. Setting $\Delta G_{\mathrm{slv}} = 0$ when the gas partial pressure and the solute's concentrations are equal to their equilibrium values $P_A$ and $C_A$, respectively, we obtain
\begin{equation}
  P_A = k_H C_A
  \label{eq:Henrys-law}
\end{equation}
where
\begin{equation}
  k_H = k_B T \frac{z_{A(\mathrm{g})}}{z_{A(v)}} \, .
  \label{eq:Henrys-constant}
\end{equation}
The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{nicholls2008predicting,mobley2014freesolv}
\begin{equation}
\Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = k_B T \ln \frac{k_H}{k_B T} \, .
\label{eq:DG-solvation-excess}
\end{equation}
Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent. This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}

@@

Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields\cite{widom1982potential} 
\begin{equation}
  \Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \langle e^{-\beta \Delta U} \rangle_0 = k_B T \ln \langle e^{\beta \Delta U} \rangle_1
  \label{eq:zratio-solv-dgexc}
\end{equation}
The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols discussed in the following sections.

@@

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}. Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of solvation or a free energy of binding. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimized structure of a solute can be generated with a simple tool such as Open Babel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced different results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species. Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested. Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units. A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).

@@

Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.

\paragraph{Handling the standard state in free energy calculations.} \label{sec:ABFE-standard-state}
As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward. Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations. These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions. Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.

@@

\begin{figure}
    % Figure to be updated
    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
    \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g. via the path shown as the dashed line. In contrast, in scenario \textbf{B}, the free energy barrier extending all alchemical states prevents the system from sampling both metastable states at $\lambda=0$. In scenario \textbf{C}, the free energy basin extending in both alchemical and configurational directions can trap the system.}
    \label{fig:configurational_sampling}
\end{figure}

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicas to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
If using a large number of replicas these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

\paragraph{Conformational sampling challenges can affect both solvation and binding free energies}
While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial. Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]
    {paper/figures_atm/atm_scheme_abfe_rbfe.pdf}
    \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations. In (A), the receptor (blue) is shown with the ligand in bulk solvent being translated to the active site along a vector $+h$. In (B), the RBFE setup is shown with two ligands: one in the binding site and the other in bulk solvent. A vector $h$ separates the ligands, and the perturbation energy is evaluated by virtually swapping their positions at each MD step.}
    \label{fig:atm-illustration}
\end{figure}

@@

The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \label{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}. One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.  The analytical relationships Eq.~(\ref{eq:ATM-softplus})--(\ref{eq:end-of-softplus}) between the binding energy and the perturbation energy make this step straightforward with ATM. In contrast, conventional free energy perturbation methods that employ soft-core pair potentials lack analytical perturbation energy functions and require a more complex process involving calling the molecular dynamics engine to rescore simulation trajectories. The more straightforward free energy estimation process afforded by ATM saves computing and storage costs, and, since it is based solely on binding energies, allows for changing the perturbation energy function $w(b;\vec{\lambda})$ without running new simulations. The simplification of the free energy estimation process fits the general ATM philosophy of streamlining and generalizing alchemical binding free energy calculations as much as possible.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]
    {paper/figures_atm/atm_alchemical_intermediate.pdf}
    \caption{Illustration of ATM's thermodynamic path. At the thermodynamic intermediate (top, center), the ligand interacts at half the strength with the receptor and half strength with the solvent. The unbound state (bottom, right) is converted to the alchemical intermediate during Leg 2. Similarly, the bound state (bottom, left) is converted to the alchemical intermediate during Leg 1. The binding free energy is recovered by taking the difference between the free energies of Leg 2 and Leg 1. The receptor is depicted in blue, and the ligand is shown in gray. The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.}
    \label{fig:atm_alchemical_intermediate}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{paper/figures_atm/atm_lig_alignment.pdf}
    \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}. Atoms \textbf{$1_A$} and \textbf{$1_B$} are selected as the first reference points, positioned near the center of the rigid common core. These atoms define the origin of the coordinate frames, ensuring that both ligands remain at a fixed distance. Atoms \textbf{$2_A$} and \textbf{$2_B$} are chosen from within the common rigid core, positioned as far as possible from the first reference atom. The axis defined by $2_A$ and $2_B$ is kept approximately collinear. Finally, atoms \textbf{$3_A$} and \textbf{$3_B$} are chosen to maintain the planar alignment of the rigid cores, preserving the roll of the common structural framework. Reference atoms are preferably not selected from flexible regions of the ligands, as changes in these regions could alter the relative orientation between the ligands.
    }
    \label{fig:atm_lig_alignment}
\end{figure}

@@

Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\label{sec:relative-fe-protocol}). In particular, as the work of~\cite{Fleck_2021} highlighted, valence terms for dummy atoms for dummy atoms must be treated far more carefully than previously appreciated in order for dummy atoms' contributions to the free energy to rigorously cancel. Less careful treatments may leave dummy atom terms only approximately canceling, or, worse, introducing uncharacterized errors in relative binding free energies. Indeed, such issues might have impeded previous attempts at reproducing free energy calculations across simulation engines~\cite{loeffler2018reproducibility}.

>>>>

Next, we outline the application of the chemical potential statistical mechanics formalism for modeling key biophysical processes: solvation, solvent partitioning, and non-covalent molecular binding. Generally, turning statistical mechanics identities into computable quantities by molecular simulations involves expressing ratios of partition functions in terms of ensemble averages. In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe}, \ref{subsec:numerical-implementations},  and \ref{subsec:estimators} rather than directly. Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice. As in the previous section, we use the solvent potential of mean force to lighten the notation when possible. The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).

@@

The standard molar Gibbs free energy of the solvation of a gas species $A$ in a solvent $v$
\begin{equation}
A(\mathrm{g})  \leftrightharpoons A(v)
\end{equation}
is
\begin{equation}
  \Delta G^\circ_{\mathrm{slv}} = \mu^\circ_{A(v)}  - \mu^\circ_{A(g)} = -k_B T \ln \frac{P^\circ}{C^\circ k_B T} - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}}
  \label{eq:DG0-solvation-def}
\end{equation}
where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel. The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases. The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below). Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.\cite{thompson2004new}

However, equilibrium experimental gas solvation data is most often available in terms of Henry's constants $k_H$. Setting $\Delta G_{\mathrm{slv}} = 0$ when the gas partial pressure and the solute's concentrations are equal to their equilibrium values $P_A$ and $C_A$, respectively, we obtain
\begin{equation}
  P_A = k_H C_A
  \label{eq:Henrys-law}
\end{equation}
where
\begin{equation}
  k_H = k_B T \frac{z_{A(\mathrm{g})}}{z_{A(v)}} \, .
  \label{eq:Henrys-constant}
\end{equation}
The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
\begin{equation}
\Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = k_B T \ln \frac{k_H}{k_B T} \, .
\label{eq:DG-solvation-excess}
\end{equation}
Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent. This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}

@@

Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields\cite{widom1982potential} 
\begin{equation}
  \Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \langle e^{-\beta \Delta U} \rangle_0 = k_B T \ln \langle e^{\beta \Delta U} \rangle_1
  \label{eq:zratio-solv-dgexc}
\end{equation}
The objective of alchemical solvation free energy calculations is to implement Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols and alchemical intermediate states as  discussed in section \ref{subsec:numerical-implementations}.

@@

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025}. Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of hydration or a free energy of binding. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

%this was moved to the theory section, keeping it for now for reference
\begin{comment}
An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species. Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested. Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units. A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).

@@

Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
\end{comment}

As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward. Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations. These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions. Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.

@@

\begin{figure}
    % Figure to be updated
    \includegraphics[width=0.88\columnwidth]{figures/fig_configurational_sampling/FES_scenarios_path.pdf}
    \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g. via the path shown as the dashed line. In contrast, in scenario \textbf{B}, the free energy barrier extending all alchemical states prevents the system from sampling both metastable states at $\lambda=0$. In scenario \textbf{C}, the free energy basin extending in both alchemical and configurational directions can trap the system.}
    \label{fig:configurational_sampling}
\end{figure}

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

\paragraph{Conformational sampling can also solvation and binding free energies}
While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial. Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]
    {figures_atm/atm_scheme_abfe_rbfe.pdf}
    \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations. In (A), the receptor (blue) is shown with the ligand in bulk solvent being translated to the active site along a vector $+h$. In (B), the RBFE setup is shown with two ligands: one in the binding site and the other in bulk solvent. A vector $h$ separates the ligands, and the perturbation energy is evaluated by virtually swapping their positions at each MD step.}
    \label{fig:atm-illustration}
\end{figure}

@@

The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \ref{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}. One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.  The analytical relationships Eq.~(\ref{eq:ATM-softplus})--(\ref{eq:end-of-softplus}) between the binding energy and the perturbation energy make this step straightforward with ATM. In contrast, conventional free energy perturbation methods that employ soft-core pair potentials lack analytical perturbation energy functions and require a more complex process involving calling the molecular dynamics engine to rescore simulation trajectories. The more straightforward free energy estimation process afforded by ATM saves computing and storage costs, and, since it is based solely on binding energies, allows for changing the perturbation energy function $w(b;\vec{\lambda})$ without running new simulations. The simplification of the free energy estimation process fits the general ATM philosophy of streamlining and generalizing alchemical binding free energy calculations as much as possible.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]
    {figures_atm/atm_alchemical_intermediate.pdf}
    \caption{Illustration of ATM's thermodynamic path. At the thermodynamic intermediate (top, center), the ligand interacts at half the strength with the receptor and half strength with the solvent. The unbound state (bottom, right) is converted to the alchemical intermediate during Leg 2. Similarly, the bound state (bottom, left) is converted to the alchemical intermediate during Leg 1. The binding free energy is recovered by taking the difference between the free energies of Leg 2 and Leg 1. The receptor is depicted in blue, and the ligand is shown in gray. The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent. \egnote{To do: reverse the directions of the arrows}}
    \label{fig:atm_alchemical_intermediate}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures_atm/atm_lig_alignment.pdf}
    \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}. Atoms \textbf{$1_A$} and \textbf{$1_B$} are selected as the first reference points, positioned near the center of the rigid common core. These atoms define the origin of the coordinate frames, ensuring that both ligands remain at a fixed distance. Atoms \textbf{$2_A$} and \textbf{$2_B$} are chosen from within the common rigid core, positioned as far as possible from the first reference atom. The axis defined by $2_A$ and $2_B$ is kept approximately collinear. Finally, atoms \textbf{$3_A$} and \textbf{$3_B$} are chosen to maintain the planar alignment of the rigid cores, preserving the roll of the common structural framework. Reference atoms are preferably not selected from flexible regions of the ligands, as changes in these regions could alter the relative orientation between the ligands.
    }
    \label{fig:atm_lig_alignment}
\end{figure}

@@

Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\ref{sec:relative-fe-protocol}). In particular, as the work of~\cite{Fleck_2021} highlighted, valence terms for dummy atoms for dummy atoms must be treated far more carefully than previously appreciated in order for dummy atoms' contributions to the free energy to rigorously cancel. Less careful treatments may leave dummy atom terms only approximately canceling, or, worse, introducing uncharacterized errors in relative binding free energies. Indeed, such issues might have impeded previous attempts at reproducing free energy calculations across simulation engines~\cite{loeffler2018reproducibility}.
